---
title: "Analyzing Sleep Data and Trends"
author: "Veronica Jenkins, Anika Deshpande, Katie Clayton"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    code folding: TRUE
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The two data sets we have chosen to use are called "How Much Sleep Do
Americans Really Get?" and "Sleep Data." Both are from the website
Kaggle. In this project, we renamed the first data set, TimeSleep while
the second remained SleepQualityData. The TimeSleep data set was
originally created to study the sleep patterns of people among different
age groups. The attributes consist of year, period, average hours per
day sleeping, standard error, type of days, age group, activity and sex.
We chose not to use all the attributes, as they did not relate to our
research questions. The attributes we did use throughout this project
from this data set are Average Hours Per Day Sleeping, Type of Days, and
Age Group. The SleepQuality data set was gathered from the Sleep Cycle
iOS app between 2014-2018. The purpose of this gathered data was to
compare the different ways people sleep, their length of sleep and how
outside factors affected how much/how well they slept. There were
multiple attributes included in this set, but we pulled a few we thought
could contribute best to our project. The attributes we used are Hours
of Sleep, Sleep Quality, and Movements Per Hour.

```{r LoadPackages}
#CamelCase
library(dplyr)
library(ggplot2)   
library(readxl)
library(tidyverse)
library(readr)
library(tidyr)
library(esquisse)

TimeSleep <- read_excel(
  path = "~/Desktop/TimeSleep.xlsx",
  skip = 1)

SleepQualityData <- read_excel("~/Desktop/SleepQualityData.xlsx", skip = 1)
```

```{r TimeSleep}


#rename columns 
colnames(TimeSleep)<- c('Index','Year','Period','AverageHoursPerDaySleeping','Standard Error','Type of Days','Age Group','Activity','Sex')

#Make values in "Average Hours Per Day Sleeping" coloumns 
# TimeSleep$`AverageHoursPerDaySleeping`<- as.numeric(TimeSleep$`AverageHoursPerDaySleeping`)
# 
# #removes first column 
# TimeSleep <- TimeSleep[-1,]
```

```{r SleepQuality}
#rename column names
colnames(SleepQualityData) <- c('Start', 'End', 'SleepQuality', 'Regularity', 'Mood', 'HeartRate(BPM)', 'Steps', 'AlarmMode', 'AirPressure(Pa)', 'City', 'MovementsPerHour', 'TimeInBed(Seconds)', 'TimeAsleep(Seconds)', 'TimeBeforeSleep(Seconds)', 'WindowStart', 'WindowStop', 'DidSnore', 'SnoreTime', 'WeatherTemperature(Celcius)', 'WeatherTypes', 'Notes')

#get rid of the first column
SleepQualityData <- SleepQualityData[-1,]

#creates new data frame of average hours of sleep
```

## Hours by Age

#### Is there change in mean hours of sleep per night depending on different age groups?

The Time Sleep data set presented two attributes that were easy to
compare: Mean Hours of Sleep Per Night, and Age Group. We chose to
incorporate this into our research because the phrase "I am so tired" is
used very frequently among many different age groups. Additionally,
different age groups are known to get more/less sleep than others, so we
wanted to explore how much the difference is. After choosing the
attributes, we made a table and were able to input the table into ggplot
to create the bar graph.

```{r WrangleHoursByAge}
#creates new data frame with of average hour per day sleeping
HoursByAge<- TimeSleep%>%
  group_by(`Age Group`)%>%
  summarize(
    Mean=mean(`AverageHoursPerDaySleeping`)
  )
HoursByAge<-HoursByAge[-2,]
```

```{r HoursByAgeVisualization}
# making the visualization for HoursByAge
#load ggplot, dplyr and HoursByAge data
library(scales)
library(ggplot2)
library(dplyr)

ggplot(
  data = HoursByAge,
  mapping = aes(x = `Age Group`, weight = Mean)
) +
  geom_bar(fill = "#FFD1E8", width = 0.67, color = "black") +
  # adding titles and labels
  labs(
    x = "Age Group",
    y = "Mean Hours of Sleep",
    title = "Mean Hours of Sleep per Night Depending on Age Group"
  ) +
  theme(axis.line.x = element_line(color="black", size = .25),
        axis.line.y = element_line(color="black", size = .25)
  ) +
  scale_y_continuous(breaks = seq(0,10, by = 0.5), limits = c(0,10), expand = c(0, 0))
```

#### What Can be Shown?

The bar graphs shows six different age groups and a range of 0 to 10
mean hours of sleep. While all the age groups seem to show similar
numbers, we can conclude that the age group 15 to 24 years has the
highest mean while the age group 45 to 54 years has the lowest. If you
were to draw a curved line at the top of the bars, it would form a
slight "u" shape showing how people will start with the max mean hours
of sleep and will slowly decrease to the min and immediately will begin
to steadily increase until reaching the max once again. Although there
is not a huge difference, it is interesting to see how the data displays
as opposed to what people may originally assume.

## Sleep Quality by Hours of Sleep

#### How is sleep quality affected by hours of sleep per night?

For this question, we pulled data from the Sleep Quality Data set
imported from excel. We selected the attributes Sleep Quality and Hours
of Sleep. The attribute Sleep Quality is formatted as a percentage and
is displayed in decimal format for the graph. Using the filtered data of
the selected column, we displayed the data in a scatter plot with a line
of best fit.

```{r SleepQualityByHoursTidyData}
#TidyData

#Create a dataframe with only the necessary columns
SleepQuality <- data.frame(SleepQualityData$'Start', SleepQualityData$'End', SleepQualityData$'SleepQuality')
#rename the columns


colnames(SleepQuality) <- c('Start', 'End', 'SleepQuality')

#Create a column calculating the difference in timestamps to get hours of sleep
SleepQuality <- transform(
  SleepQuality, HoursOfSleep= as.numeric(difftime(as.POSIXct(SleepQualityData$`End`), as.POSIXct(SleepQualityData$`Start`)), units = "hours"))

#round hours of sleep
SleepQuality$`HoursOfSleep` <- round(SleepQuality$`HoursOfSleep`, digits = 2)

#Convert sleep quality to numeric
SleepQualityData$`SleepQuality`<- as.numeric(SleepQualityData$`SleepQuality`)

#load a package to convert sleep quality to percent
library(scales)
library(formattable)
SleepQuality$`SleepQuality` <- formattable::percent(SleepQualityData$`SleepQuality`)

#Create a dataframe with the two neccessary columns and rename the columns
SleepQualityByHours<-data.frame(SleepQuality$`HoursOfSleep`, SleepQuality$`SleepQuality`)
colnames(SleepQualityByHours) <- c('HoursOfSleep', 'SleepQualityPercent')
```

```{r SleepQualityByHoursVisualization}
#use ggplot to create a line graph using the data created 
ggplot(
  data = SleepQualityByHours,
  mapping = aes(x = HoursOfSleep, y = SleepQualityPercent)
  ) +
  #customize point size, color, and shape
  geom_point(shape = "circle", size = 1.0, colour = "#BF1B6D") +
  #create a line of best fit and get rid of the spanning color
  geom_smooth(
    span = 0.75,
    se = FALSE,
    colour = "black",
    method = "loess",
    formula = y~x) +
  #create titles and axis labels
  labs(
    x = "Hours of Sleep",
    y = "Sleep Quality Percent",
    title = "Hours of Sleep Vs. Sleep Quality"
  ) +
  #set the graph appearance and fix scaling issues
  theme_minimal() +
  scale_x_continuous(breaks = seq(0,14, by = 1), limits = c(0,14), expand = c(0, 0))
```

#### What is Shown?

The graph shows that most people are logging anywhere between 5 to 10
hours a night. The relationship between hours of sleep and sleep quality
show an exponential curve. There is a clear increase and trend showing
the more sleep a person gets, the better their sleep was however around
10 hours, the line plateaus, meaning sleep quality is likely to no
longer improve. Overall, the graph displays 10 hours of sleep maximizes
sleep quality.

## Hours of Sleep by Day Classification

#### How is the amount of sleep affected by what type of day it is?

For this research question, we pulled data from the TimeSleep data set.
We wanted to compare statistical averages such as the min, 1st quartile,
median, 3rd quartile, max, and outliers of hours of sleep per night
grouped by what type of day it was. For example, a day classification
would be non-holiday weekdays, weekend days and holidays, and all days
(all types). Since weekends and holidays are presumed to be more relaxed
for people, we thought it would be interesting to explore how much of a
difference there actually is. This question will be explored through a
box-plot and summary table to display the statistical averages
mentioned.

#### Box-Plot Visualization

```{r HoursByDayType}
#create an object for filtered data
HoursByDay <- TimeSleep %>%
  #organize all of the same day types together
  group_by(`Type of Days`) %>%
  #select the two columns needed: day type and average hours asleep 
  select(`Type of Days`, `AverageHoursPerDaySleeping`)
```

```{r HoursByDayTypeVisualization}
#use ggplot to create a boxplot
ggplot(
  data = HoursByDay,
  mapping = aes(x = `Type of Days`, y = AverageHoursPerDaySleeping)
  ) +
  #assign a color to fill the box plot
  geom_boxplot(fill = "#FFD1E8") +
  #create titles and axis labels
  labs(
    x = "Day Classification",
    y = "Hours of Sleep",
    title = "Hours of Sleep by Day Classification"
  ) +
  #set graph appearance and fix axis scaling
  theme_minimal() +
  ylim(7.5, 11)
```

#### What is Shown?

As expected, the box-plot displays that Non-holiday weekdays have the
lowest median hours of sleep and Weekend days and holidays having the
highest median hours of sleep. There are no outliers in non-holiday
weekdays showing consistent data within this category. There are however
many outliers in the Weekend days and holidays group showing a higher
number of hours of sleep. Going back to the proposed research question
asking if people really get significantly more sleep on weekends and
holidays or if it is just assumed, it is safe to conclude that the
median is roughly one hour higher which is a relatively significant
difference. In comparison, the overall (All days) median lies at roughly
8.6. These facts overall suggest weekend days and holidays do have a
significant increase in amount of sleep, however the difference in sleep
is not very large.

#### Summary Table

```{r HoursByDaySummaryTable}
library(readxl)
library(kableExtra)
library(janitor)



colnames(TimeSleep)<- c('Index','Year','Period','AverageHoursPerDaySleeping','Standard Error','Type of Days','Age Group','Activity','Sex')

TimeSleep <- TimeSleep[-1,]
DaysVsSleep <- TimeSleep %>%
  group_by(`Type of Days`) %>%
  select(`Type of Days`, `AverageHoursPerDaySleeping`) %>%
  summarize(
    across(
      .cols = where(is.numeric),
      .fns = list(
        min = ~min(`AverageHoursPerDaySleeping`, na.rm = TRUE),
        Q1 = ~quantile(`AverageHoursPerDaySleeping`, probs = 0.25, na.rm = TRUE),
        median = ~median(`AverageHoursPerDaySleeping`, na.rm = TRUE),
        Q3 = ~quantile(`AverageHoursPerDaySleeping`, probs = 0.75, na.rm = TRUE),
        max = ~max(`AverageHoursPerDaySleeping`, na.rm = TRUE),
        sam = ~mean(`AverageHoursPerDaySleeping`, na.rm = TRUE),
        sasd = ~sd(`AverageHoursPerDaySleeping`, na.rm = TRUE)
      ),
      #round the values to avoid unecessarily large amount of decimal spaces
      round(digits = 2)
    ),
    #get count and format it so that big numbers are separated with commas
    count = format(n(), big.mark = ","),
  )
#rename the columns 
colnames(DaysVsSleep) <- c("Day Classification", "Min", "1st Quartile", "Median", "3rd Quartile", "Max", "Arithmetic Mean", "Arithmetic Standard Deviation", "Count")
#use kable to tidy the table (add caption, add grid, align columns) 
DaysVsSleep %>%
  kable(
    caption = "Statistics on Average Hours Spent Sleeping and Type of Day",
    booktabs = TRUE,
    align = c("l", rep("c", 6))
  ) %>%
  #separate rows through striping the colors
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "condensed"),
    font_size = 16
  ) %>%
  kableExtra::kable_classic()
```

#### What is Shown?

As a visual aid to the box-plot, we have created a summary table of the
similar statistics shown on the box-plot in a more readable format. The
addition of the count column shows the amount of rows used to calculate
each group. Each group has a count of 315 showing an equal amount of
data being used to compute the statistics for each category. There is
also the addition of the arithmetic mean which can be used as another
form of the median to display an average. The final and last addition is
the arithmetic standard deviation. The summary table shows the same
information as the box-plot in which people are more likely to sleep
more on weekends and holidays and see a significant difference in sleep
but not by a very large amount.

## Sleep Quality by Movements During Sleep

#### How do number of movements per hour affect sleep quality of an individual?

From the Sleep Quality data set, we proposed a question to see if there
is any correlation between these two attributes. The two attributes used
from the data set were \`Sleep Quality\` along the y axis and
\`Movements Per Hour\` along the x axis. We also included a line of best
fit after using ggplot to create a scatter plot showing correlation.

```{r MovementsVSleepQuality }

#convert column data to numeric
SleepQualityData$`MovementsPerHour`<- as.numeric(SleepQualityData$`MovementsPerHour`)
SleepQualityData$`SleepQuality`<- as.numeric(SleepQualityData$`SleepQuality`)

#filter data
movementsVsTime <- SleepQualityData %>%
  select(MovementsPerHour, SleepQuality)

#subset data
movementsVsTime <- movementsVsTime[0:250, ]

#use ggplot to create a line graph. Select the data and x&y axis
ggplot(
  data = movementsVsTime,
  mapping = aes(x = MovementsPerHour, y = SleepQuality )
  ) +
  geom_point(shape = "circle", colour = "#BF1B6D") + #assign the points a shape and color and add a line of best fit
  geom_smooth(
    span = 0.75,
    se = FALSE,
    colour = "black",
    method = "loess", 
    formula = y ~ x
  ) +
  geom_point(shape = "circle", colour = "#BF1B6D") +
  # adding titles and labels
  labs( # adding titles and labels
    x = "Movements Per Hour",
    y = "Sleep Quality",
    title = "Movements Per Hour vs Sleep Quality"
  ) +
  theme_minimal() + #set the graph appearance
  scale_x_continuous(breaks = seq(0,150, by = 15), expand = c(0, 0))  #fix the x axis scaling
```

#### What is Shown?

As shown on the scatter plot there is a correlation between these two
attributes. As most would assume, it is a negative correlation as well.
Our research proves that as the number of movements per hour increase,
the sleep quality of an individual decreases. As shown on the scatter
plot there is a correlation between these two attributes. As most would
assume, it is a negative correlation as well. Our research proves that
as the number of movements per hour increase, the sleep quality of an
individual decreases. Most of the movements range between 25 and 75 with
outliers at 0 and 125. As for sleep quality, most of the points range
from around 6.5 to 1.0.

## Conclusion

From our two data sets were able to gain a lot of insight about various
factors that impact sleep. We were able to assess the amount of sleep
that various age groups get each night. Our data also highlighted the
positive relationship between the number of hours of sleep and sleep
quality. This indicated that people of all age groups could benefit from
sleeping more. Another characteristic impacting sleep quality was number
of movements per hour. Our data showed a negative correlation between
movements per hour and sleep quality indicating that more movement leads
to lower quality sleep. We also found that typically people are getting
more sleep on weekends and holidays than they are on a normal weekday.
These visualizations could be used to help develop better ways to
achieve quality sleep.
