---
title: "Final Project"
author: "Veronica Jenkins, Anika Deshpande, Katie Clayton"
date: "`r Sys.Date()`"
output: html_document  
editor_options: 
  markdown: 
    wrap: 72
---

```{r LoadPackages}

library(dplyr)
library(ggplot2)
library(readxl)
library(tidyverse)
library(readr)
library(tidyr)

TimeSleep <- read_excel("~/Desktop/TimeSleep.xlsx")
SleepQualityData <- read_excel("~/Desktop/SleepQualityData.xlsx")
View(SleepQualityData)
```

```{r TimeSleep}
#load data 
TimeSleep <- read_excel("~/Desktop/TimeSleep.xlsx")

#rename columns 
colnames(TimeSleep)<- c('Index','Year','Period','Average Hours Per Day Sleeping','Standard Error','Type of Days','Age Group','Activity','Sex')

#Make values in "Average Hours Per Day Sleeping" coloumns 
TimeSleep$`AverageHoursPerDaySleeping`<- as.numeric(TimeSleep$'Average Hours Per Day Sleeping')

```{r TimeSleep}
#rename columns
colnames(TimeSleep)<- c('Index','Year','Period','AverageHoursPerDaySleeping','Standard Error','Type of Days','Age Group','Activity','Sex')
TimeSleep <- TimeSleep[-1,]
#Make values in "Average Hours Per Day Sleeping" coloumns 
TimeSleep$`AverageHoursPerDaySleeping`<- as.numeric(TimeSleep$`AverageHoursPerDaySleeping`)

#removes first column 
TimeSleep <- TimeSleep[-1,]
```

```{r WrangleHoursByAge}
TimeSleep$AverageHoursPerDaySleeping<- as.numeric(as.character(TimeSleep$'AverageHoursPerDaySleeping'))


```

```{r }
#rename column names
colnames(SleepQualityData) <- c('Start', 'End', 'SleepQuality', 'Regularity', 'Mood', 'HeartRate(BPM)', 'Steps', 'AlarmMode', 'AirPressure(Pa)', 'City', 'MovementsPerHour', 'TimeInBed(Seconds)', 'TimeBeforeSleep(Seconds)', 'WindowStart', 'WindowStop', 'DidSnore', 'SnoreTime', 'WeatherTemperature(Celcius)', 'WeatherTypes', 'Notes')

#get rid of the first column
SleepQualityData <- SleepQualityData[-1]
```

```{r WrangleHoursByAge}
HoursByAge<- TimeSleep%>%
  group_by(`Age Group`)%>%
  summarize(
    Mean=mean(`AverageHoursPerDaySleeping`)
  )
HoursByAge<-HoursByAge[-2,]
```

```{r HoursByAgeVisualization}
# making the visualization for HoursByAge
#load ggplot, dplyr and HoursByAge data
library(ggplot2)
data(HoursByAge)
library(dplyr)
ggplot(
  data = HoursByAge,
  mapping = aes(x = `Age Group`, y = Mean)
) +

  geom_point(shape = "circle", size = 4L, colour = "#112446") +
  
  # adding titles and labels
  labs(
    x = "Age Group",
    y = "Mean Hours of Sleep",
    title = "Mean Hours of Sleep per Night Depending on Age Group"
  ) +
  theme_minimal()
```

```{r SleepQualityByHoursVisualization}
#

```

```{r}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
